---
title: "Daily Census"
output:
  html_vignette:
    number_sections: true
    toc: true
    dfprint: kable

vignette: >
  %\VignetteIndexEntry{Daily Census}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

# Overview

The `daily_census` function calculates the number of patients that were hospitalized at a given hospital on each day during a time period of interest. 
The `census` provides a cross-sectional count of patients that occupied a hospital bed at 8am (by default) each day. Specifically, it is calculated as the number of patients who were admitted before that time point and discharged after that time point. The function also returns a `capacity_ratio` indicating whether the number of patients on a given day was higher (>1) or lower (<1) than on a typical day, where by default, typical occupancy is based on the median census during the time period of interest. 

**Note:** Although this function can be applied to different cohorts and research contexts, the interpretation of the function outputs may vary according to the inputs users provide. For example, if users pre-filter their cohort based on certain diagnosis criteria, the `census` counts will only include those patients, and similarly, the `capacity_ratio` would not reflect a real indicator of capacity, but rather a measure of whether occupancy is above/below typical occupancy *for a certain group of patients* (also see [*section on group_var input*](#grouping) below). 

# Running `daily_census()` with default arguments

To run the `daily_census` function, users need to provide an input table `cohort` containing all encounters that are part of their cohort of interest.
Typically, the cohort is created based on the `admdad` table (or a subset thereof) in the GEMINI database. It needs to contain the columns `genc_id`, `hospital_num`, `admission_date_time`, and `discharge_date_time`.

Here is an example of how to load the relevant data and run the `daily_census` function with default settings:

```{r, eval=F}
# Load necessary libraries
library(RPostgreSQL)
library(DBI)
library(getPass)

# Establish database connection
db <- DBI::dbConnect(drv,
                     dbname = "DB_name",
                     host = "172.XX.XX.XXX",
                     port = 1234,
                     user = getPass("Enter user:"),
                     password = getPass("Enter Password:"))

# query admdad table containing cohort (could be filtered further if necessary)
admdad <- dbGetQuery(db, "SELECT genc_id, hospital_num, admission_date_time, discharge_date_time FROM admdad;") 

# Run default daily_census calculation         
census_output <- daily_census(cohort = admdad) 

head(census_output, 10)

```

> **Mock output table (for illustration purposes, not real data):**

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)

data(dummy_census, package = "Rgemini")
dummy_census <- dummy_census %>% data.table()
dummy_census[ , hospital_num := as.factor(hospital_num)]

# dummy date includes larger time period and grouping variables, let's ignore them for now
census_output <- dummy_census[as.Date(date_time) < '2017-04-01', .(census = sum(census)), by = c("date_time","hospital_num")]
census_output <- census_output[order(date_time)]
census_output[ , capacity_ratio := census / median(census), by = hospital_num]

# show output example
head(census_output,10) %>% knitr::kable(align=c('c','c','c','c'))

```

This table shows an example output for data from 5 hospitals from April 2016 - April 2017. Each row corresponds to a unique combination of `date_time` (with a reference time of 8am, by default) and hospital ID (`hospital_num`). The `census` variable indicates the count of patients that were occupying a bed at 8am each day at a given hospital. The `capacity_ratio` refers to a relative measure of bed occupancy, which by default is calculated as `census/median(census)`. 

Based on this table, users could extract further information, such as the median daily census at each hospital during the study period:

```{r, eval=T}
library(data.table) #  we are using data.table operations below, but users could also use tidyverse to analyze the function output

# Compute median census at each site
median_census <- census_output[ , .(median_census = median(census)), by = hospital_num] 

```

```{r, eval=T, echo=F}
median_census %>% knitr::kable(align=c('c','c'))
```


Additionally, users could plot `census` and `capacity_ratio` over time, separately for each hospital: 

```{r, eval=T,fig.height=5,fig.width=8,out.width = "80%", fig.align = "center"}
library(ggplot2)

# Plot census over time
ggplot(census_output, aes(x=as.Date(date_time), y=census, group = hospital_num, color=hospital_num)) +
  geom_line(linewidth=1.5,show.legend = TRUE) +
  scale_x_date(name = 'Date', breaks = seq(min(as.Date(census_output$date_time)), max(as.Date(census_output$date_time)), by="1 months"), date_labels = "%b\n%Y") +
  ggtitle("Mock figure: Daily census by hospital") + theme_classic()

```


```{r, eval=T,fig.height=5,fig.width=8,out.width = "80%", fig.align = "center"}
# Plot capacity ratio over time
ggplot(census_output, aes(x=as.Date(date_time), y=capacity_ratio, group = hospital_num, color=hospital_num)) +
  geom_line(linewidth=1.5,show.legend = TRUE) +
  scale_x_date(name = 'Date', breaks = seq(min(as.Date(census_output$date_time)), max(as.Date(census_output$date_time)), by="1 months"), date_labels = "%b\n%Y") +
  ggtitle("Mock figure: Daily capacity ratio by hospital") + theme_classic() +
  geom_hline(yintercept=1)

```

Note that `capacity_ratio` fluctuates around 1 (= typical occupancy) where values > 1 correspond to days with higher occupancy than usual.


# Optional input arguments 

## `time_period`

By default, the function will calculate the census for the whole time period that is available in the `cohort` table. Note that data availability may differ by hospital, and the function will determine data availability individually for each site. If users only want to calculate the census for a certain time period within their overall cohort, they can provide an optional `time_period` input specifying a start and end date. 

For example, to calculate the census from June 1, 2016 - Dec 31, 2016: 

```{r, eval=F}
# Compute census for specific time period
census_output <- daily_census(admdad, time_period = c("2016-06-01","2016-12-31")) 

```

Note that in this case, the `capacity_ratio` is calculated based on the typical occupancy observed during that same time period (i.e., between June-Dec 2016). 


## `scu_exclude`

By default, the total duration of each encounter's hospital stay is considered in the census counts. However, for certain projects, it may be of interest to exclude any time points where the encounter was in a special care unit (SCU), such as intensive care (ICU). This is relevant if researchers want to analyse bed occupancy at a particular medical ward (e.g., GIM) and only want to count patients who were in fact occupying a bed in that ward on a given day, while excluding any patients who were in an SCU. 

In that case, users should provide an `scu_exclude` table that contains all SCU encounters that should be excluded from the census calculation.
Note: The SCU table typically refers to the `ipscu` table in the GEMINI database, however, users may want to further filter that table by relevant `scu_unit_numbers` that should be excluded from the census counts. The function automatically removes any entries where `scu_unit_number = 99`, which refers to encounters with `"no SCU"`. Additionally, only SCU encounters with a valid `scu_admit_date_time` and `scu_discharge_date_time` can be excluded from the census. Availability of these variables is low for certain cohorts and SCU units. Therefore, users are advised to carefully inspect the SCU table to make an informed decision about whether to exclude SCU encounters, and if yes, which SCU entries to exclude. 

```{r, eval=F}
# exclude SCU encounters from census
scu <- dbGetQuery(db, "SELECT * FROM ipscu;") 

# Compute census excluding SCU
census_output <- daily_census(admdad, scu_exclude = scu) 

```


## `group_var` {#grouping}

By default, the census is calculated separately for each hospital (by `hospital_num`). Users can specify additional grouping variables to obtain patient counts (and capacity ratios) for subgroups of interest, such as different medical subservices or physicians.

Here is a simple example where census is grouped by patients' gender and age (<65 vs. 65+):

```{r, eval=F}
# Create age category
admdad$age_cat <- ifelse(admdad$age <= 65,'<=65','>65')

# Compute census by gender & age categories
census_output <- daily_census(admdad, group_var = c("gender","age_cat")) 

head(census_output,10)
```

> **Mock output table showing census grouping by age & gender categories:**

```{r, eval=T, echo=FALSE}
head(dummy_census) %>% knitr::kable(align=c('c','c'))
```

Note that `capacity_ratio` in this mock example should not be interpreted as a real "capacity" indicator, but rather, as a measure of whether on a given day there were more (>1) or less (<1) patients of a certain age/gender than on a typical day. `Capacity_ratio` is more useful for grouping variables that correspond to separate medical entities, such as different medical subservices or wards, where `capacity_ratio` can serve as an indicator of system load vs. capacity. Nevertheless, grouping variables that are based on patient characteristics (age, gender, diagnosis group, illness severity etc.) could be useful to analyse the case mix of hospitalized patients during certain periods of time. 

## `capacity_func` 

By default, `capacity_ratio` is defined as `census/median(census)`, which is calculated separately for each hospital and grouping variable (if any). That is, `capacity_ratio` refers to the daily count of patients relative to typical bed occupancy, where typical bed occupancy is defined as the median census during the time period of interest. Alternatively, users can specify other measures of central tendency to obtain typical occupancy ("`mean`", "`mode`") or estimate capacity based on the maximum occupancy ("`max`").


```{r, eval=F}
# Get occupancy relative to max capacity (estimated based on max(census))
census_output <- daily_census(admdad, capacity_func = "max") 

```


## `time_of_day` 

By default, the census is calculated at 8am each day during the study period. For example, to obtain the patient counts for April 1st, 2016 the function counts all patients with `admission_date_time <= '2015-04-01 08:00:00'` and `discharge_date_time >= '2015-04-01 08:00:00'`. Users can specify a different reference time by providing an optional `time_of_day` input: 

```{r, eval=F}
# Calculate census at 2.30pm each day
census_output <- daily_census(admdad, time_of_day = "14:30:00") 

```

## `buffer` 

For time periods that are towards the end of the data availability timeline of a given hospital, users may observe a pattern similar to this, where census counts suddenly drop at the end of the specified time period:

```{r, result = 'asis', echo=F,eval=T,fig.height=5,fig.width=8,out.width = "80%", fig.align = "center",warning=F}

census_output <- dummy_census[date_time >= '2017-04-01', .(census = sum(census)), by = c("date_time","hospital_num")]

ggplot(census_output, aes(x=as.Date(date_time), y=census, group = hospital_num, color=hospital_num)) +
  geom_line(linewidth=1.5,show.legend = TRUE) +
  scale_x_date(name = 'Date', breaks = seq(min(as.Date(census_output$date_time)), max(as.Date(census_output$date_time)), by="1 months"), date_labels = "%b\n%Y") +
  ggtitle("Mock figure: Daily census by hospital - truncation bias (no buffer period)") + theme_classic()

```

This effect is due to a truncation bias that can occur if the end of the specified time period (here April 2018) is close to the last available date in the overall cohort (e.g., the cohort only contains encounters up to April 2018 in this example). Note that cohorts are typically defined by *discharge* date. That is, if a cohort includes data from April 2017 - April 2018, it will only include encounters that were *discharged* during that time. As a result, there may be patients who were admitted during the last week of April 2018, but were not discharged prior to the end of the month. Therefore, they are not part of the cohort, and cannot be counted towards the census. 

To prevent this from biasing the `census` and `capacity_ratio` estimates, the function automatically checks for data availability at each hospital based on the min and max dates in the `cohort` input table. If data availability ends prior to (or at the same time as) the end of the time period of interest, a buffer period of 30 days is applied by default. Specifically, the last 30 days at each site will be set to `NA`. The default setting of 30 days is based on the observation that the vast majority of hospital stays are < 30 days, and therefore, we can be confident that patients who where hospitalized 30 days prior to the end of the specified time period have already been discharged (i.e., are included in the cohort). 

The default setting of `buffer = 30` will result in the following output, which removes the truncation bias shown in the previous figure by setting the last 30 days of the time period to `NA`:  

```{r, result = 'asis', echo=F,eval=T,fig.height=5,fig.width=8,out.width = "80%", fig.align = "center",warning=F}

census_output[census_output[,tail(.I,30),by=hospital_num]$V1]$census <- NA

ggplot(census_output, aes(x=as.Date(date_time), y=census, group = hospital_num, color=hospital_num)) +
  geom_line(linewidth=1.5,show.legend = TRUE) +
  scale_x_date(name = 'Date', breaks = seq(min(as.Date(census_output$date_time)), max(as.Date(census_output$date_time)), by="1 months"), date_labels = "%b\n%Y") +
  ggtitle("Mock figure: Daily census by hospital - buffer period applied to last 30 days") + theme_classic()

```

Users can specify other buffer periods (in full days) based on their cohort (and the typical length of stay observed in that cohort). For example, to set the buffer period to 10 days, run the following code:

```{r, eval=F}
# specify buffer of 10 days at end of time period
census_output <- daily_census(admdad, buffer = 10) 

```

Note: In this example, if users specify a `time_period` input that ends 10 days earlier than the latest available data, the buffer period will be ignored and all available data will be used to calculate census for the time period of interest. In other words, the actual buffer period that is applied within the function depends on a combination of the specified time period, data availability per site, and the buffer argument in order to ensure that the maximum amount of available data are used in the census calculation. 


# Calculating customized measures of capacity

In addition to the flexibility provided by the function itself, users may want to obtain additional capacity indicators that are currently not supported by the function. For example, researchers may want to calculate `capacity_ratio` relative to the typical occupancy on a **year-over-year basis**. The function calculates `capacity_ratio` by estimating typical (or max) capacity based on census numbers throughout the whole time period. However, the census may vary on a year-by-year basis. In that case, users could either run the `daily_census()` function separately for each study year, or they could run the function on the whole time period and calculate a year-over-year `capacity_ratio` based on the raw `census` output provided by the function, e.g.:  

```{r, eval=F}
# calculate capacity_ratio based on median census *per year*
census_output[ , year := year(census_output$date_time)]
census_output[ , capacity_ratio_yoy := census/median(census), by = c("hospital_num","year")]

```

Similarly, users could define other indicators of typical occupancy that are currently not supported by the function (e.g., trimmed means).
