---
title: "Neighbourhood-level socioeconomic (SES) variables"
output:
  html_vignette:
    number_sections: true
    toc: true
    toc_depth: 3
    dfprint: kable

vignette: >
  %\VignetteIndexEntry{Neighbourhood-level socioeconomic (SES) variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE, 
  eval = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  collapse = TRUE,
  cache = FALSE, 
  tidy = FALSE,
  fig.height = 6, 
  fig.width = 8,
  fig.align = "left"
  
)

library(data.table)
library(dplyr)



```

------------------------------------------------------------------------

# Introduction

Purpose

-   This vignette introduces commonly used neighborhood-level socioeconomic (SES) variables at GEMINI. After reading this vignette, you'll gain necessary understanding of variable concepts and corresponding application recommendations.

Generic usage of SES variables

-   In descriptive/exploratory analysis, they are reported as baseline characteristics
-   In modelling, they usually act as exposures, confounders, or mediators

------------------------------------------------------------------------

# Data source

## [Statistics Canada Census](https://www12.statcan.gc.ca/census-recensement/2021/ref/98-304/2021001/chap1-eng.cfm)

The Census of Population provides a detailed and comprehensive statistical portrait of Canada. It’s collected by Statistics Canada conducts once every 5 years, aiming to provide population and dwelling counts for communities of all sizes across Canada.

### Ontario Marginalization Index (ON-Marg)

Concept

-   On-Marg is a neighborhood-level index showing marginalization differences between areas. It’s created using a portion of Statistics Canada Census variables.

How ON-Marg index is generated

-   Based on theoretical framework, 42 Census variables are selected.
-   After principal component factor analysis, 18 Census variables are used to create 4 ON-Marg dimensions.
-   Each dimension is available in quintiles – Q1 represents the least marginalization, while Q5 represents the most marginalization.

References

-   [ON-Marg guide 2016](https://www.publichealthontario.ca/-/media/documents/U/2018/userguide-on-marg.pdf)
-   [ON-Marg guide 2021](https://www.publichealthontario.ca/-/media/Event-Presentations/2023/09/ontario-marginalization-index-updates-products.pdf?rev=07baae2569164c17abaa18464075aa20&sc_lang=en)

# Dissemination area

Concept

-   Dissemination area is ['a small area composed of one or more neighbouring dissemination blocks and is the smallest standard geographic area for which all census data are disseminated.'](https://www150.statcan.gc.ca/n1/en/catalogue/92-169-X)

How to choose between 2016 or 2021 dissemination area

-   We recommend you to choose based on median year of your cohort, eg. if your cohort ranges from 2015 to 2020, median is 2017.5, 2016 dissemination area is preferred

# Variables ouputs from *nbhd_ses_variables* function

## Linkage variables

```{r echo=FALSE}

knitr::kable (data.table(`variables` = c('genc_id', 'da16uid', 'da21uid'),
           `definitions` = c('encounter ID', '2016 dissemination area ID', '2021 dissemination area ID')
          ) 
)
			

```

## ON-Marg index

### Quintiles

```{r echo=FALSE}

knitr::kable (data.table(`2016 variables` = c('instability_q_da16', 'deprivation_q_da16', 'dependency_q_da16', 'ethniccon_q_da16'),
           `2016 definitions` = c('1=least unstable, 5=most unstable', '1=least deprived, 5=most deprived', '1=least dependent, 5=most dependent', '1=least ethnically concentrated, 5=most ethnically concentrated'),
           `2021 variables`= c('households_dwellings_q_DA21', 'material_resources_q_DA21', 'age_labourforce_q_DA21', 'racialized_NC_pop_q_DA21'),
           `2021 definitions`= c('1=low household and dwelling-related marginalization, 5=high household and dwelling-related marginalization', '1=low material resources-related marginalization, 5=high material resources-related marginalization', '1=low age and labour force-related marginalization, 5=high age and labour force-related marginalization', '1=low racialized and newcomer population-related marginalization, 5=high racialized and newcomer population-related marginalization')
          ) 
)
			

```

### Principal components

Each index is a principal component factor score. The higher the value, the more vulnerability it presents.

```{r echo=FALSE}

knitr::kable (data.table(`2016 variables` = c('instability_da16', 'deprivation_da16', 'dependency_da16', 'ethniccon_da16'),
           
           `2021 variables`= c('households_dwellings_DA21', 'material_resources_DA21', 'age_labourforce_DA21', 'racialized_NC_pop_DA21')
          
          )
)
			

```

### Additional conceptual notes

Why name change from year 2016 to 2021

-   To avoid deficit-based language and better reflect the census measures associated with each dimension

Missingness interpretation

-   Null values: Dissemination area missing itself or no linked ON-Marg index for dissemination area

Use recommendation

-   Quintile for interpretation (categorical)
-   Principal component for modelling (continuous)

Summary score for all dimensions

-   Summary score is only recommended if all dimensions are in the same direction.
-   Summary Score = (instability quintile + deprivation quintile + dependency quintile + ethniccon quintile) / 4
-   If \>1 dimensions are in opposite direction, it’s recommended to report those dimensions separately. For example, instability and deprivation need to be reported separately if instability is a protective factor of outcome, while deprivation is a risk factor of outcome.

## Income

Variable names are the same in 2016 and 2021

```{r echo=FALSE}

knitr::kable (data.table(`variables` = c('qnatippe', 'qnbtippe', 'qaatippe', 'qabtippe', 'atippe', 'btippe'),
          
           `definitions` = c('Neighbourhood income quintile after tax (national)', 'Neighbourhood income quintile before tax (national)', 'Neighbourhood income quintile after tax (in CMA/CA)', 'Neighbourhood income quintile before tax (in CMA/CA)', 'Neighbourhood after tax income per single person equivalent (ATIPPE)', 'Neighbourhood before tax income per single person equivalent (BTIPPE)')
          
          )
)


```

[Use recommendation](https://www.cihi.ca/sites/default/files/document/toolkit-area-level-measurement-pccf-en.pdf)

-   To report inequities at provincial level, it’s recommended to use `qaatippe` or `qabtippe`, quintiles constructed separately for each census metropolitan area (CMA), census agglomeration (CA) or residual area within each province.
-   To report inequities at national level, it’s recommended to use `qnatippe` or `qnbtippe`, quintiles calculated based on national income distribution
-   To model, it's recommended to use `atippe` or `btippe`, since they are continuous

Missingness interpretation

-   Null values or 9 or 999: Unknown or missing income

## Visible minority

```{r echo=FALSE}

knitr::kable (data.table(`2016 variables` = c('c16_vismin_pct'),
           
           `2021 variables`= c('c21_vismin_pct'),
           `definitions` = c('Percentage of visible minority') )
)
			

```

[Concept](https://www12.statcan.gc.ca/census-recensement/2016/ref/dict/pop127-eng.cfm)

-   Visible minorities are defined as "persons, other than Aboriginal peoples, who are non-Caucasian in race or non-white in colour"

Visible minority group includes

-   South Asian
-   Chinese
-   Black
-   Filipino
-   Latin American
-   Arab
-   Southeast Asian
-   West Asian
-   Korean
-   Japanese
-   Visible minority, n.i.e. (not included elsewhere)
-   Multiple visible minorities
-   [Aboriginal persons are included in the Not a visible minority group](https://www23.statcan.gc.ca/imdb/p3VD.pl?Function=getVD&TVD=257518&CVD=257519&CPV=2&CST=19102015&CLV=2&MLV=2)

Calculations

-   `c16_vismin_pct` = `c16_vismin` / (`c16_vismin` + `c16_vismin_not`) x100
-   `c21_vismin_pct` = `c21_vismin` / (`c21_vismin` + `c21_vismin_not`) x100

Missingness interpretation

-   Null values: missing dissemination area or missing visible minority estimates

## Immigrant status

```{r echo=FALSE}

knitr::kable (data.table(`2016 variables` = c('c16_immsta_pct'),
           
           `2021 variables`= c('c21_immsta_pct'),
           `definitions` = c('Percentage of Immigrant status') 
          
          )
)
			

```

[Concept](https://www12.statcan.gc.ca/census-recensement/2021/dp-pd/prof/details/page.cfm?LANG=E&GENDERlist=1,2,3&STATISTIClist=4&HEADERlist=23&SearchText=Canada&DGUIDlist=2021A000011124)

-   'Immigrant status refers to whether the person is a non-immigrant, an immigrant or a non-permanent resident.'

Calculations

-   `c16_immsta_pct` = `c16_immsta_imm` / `c16_immsta` \* 100
-   `c21_immsta_pct` = `c21_immsta_imm` / `c21_immsta` \* 100

Missingness interpretation

-   Null values: missing dissemination area or missing immigrant status estimates

## Post-secondary education

```{r echo=FALSE}

knitr::kable (data.table(`2016 variables` = c('c16_ed_15over_postsec_pct', 'c16_ed_25to64_postsec_pct'),
           
           `2021 variables`= c('c21_ed_15over_postsec_pct', 'c21_ed_25to64_postsec_pct'),
                     `definitions` = c('Percentage of post-secondary education (>15 years old)', 'Percentage of post-secondary education (25-64 years old)') 
          )
)
			

```

[Concept](https://www150.statcan.gc.ca/n1/pub/81-004-x/2010001/def/posteducation-educpost-eng.htm)

-   'Postsecondary education refers to those whose highest level of educational attainment is an apprenticeship or trades certificate or diploma (including 'centres de formation professionnelle'); college, CEGEP or other non-university certificate or diploma; university certificate or diploma below bachelor level; or a university degree (bachelor's degree; university certificate or diploma above bachelor level; degree in medicine, dentistry, veterinary medicine or optometry; master's degree; earned doctorate).'

Calculations

-   `c16_ed_15over_postsec_pct` = `c16_ed_15over_postsec` / `c16_ed_15over` \* 100
-   `c16_ed_25to64_postsec_pct` = `c16_ed_25to64_postsec` / `c16_ed_25to64` \* 100
-   `c21_ed_15over_postsec_pct` = `c21_ed_15over_postsec` / `c21_ed_15over` \* 100
-   `c21_ed_25to64_postsec_pct` = `c21_ed_25to64_postsec` / `c21_ed_25to64` \* 100

Missingness interpretation

-   Null values: missing dissemination area or missing post-secondary education estimates

<!-- remove border around images -->

```{=html}
<style>
img {
border: 0;
}
</style>
```
