---
title: "Plotting Functions - Themes & Colors"
output:
  html_vignette:
    number_sections: true
    toc: true
    toc_depth: 3
    dfprint: kable

vignette: >
  %\VignetteIndexEntry{Plotting Functions - Themes & Colors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE, 
  eval = TRUE, 
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  collapse = TRUE,
  cache = FALSE, 
  tidy = FALSE,
  fig.height = 3.5, 
  fig.width = 3.5,
  fig.align = "left"
  
)

## uncomment this when trying to knit vignette locally
# library(ggpubr)
# library(ggthemes)
# library(ggpubr)
# library(scales)
# source("H:/GitHub/Rgemini/R/plot_theme.R")

options(knitr.kable.NA = '')

```


*  *  *  *

# Introduction

This vignette introduces a standardized plotting theme ([*plot_theme()*](#plot_theme)) and color palettes ([*gemini_colors()*](#gemini_colors)) that users can apply to any ggplot figures they create in their analyses. The goal of these functions is to simplify the process of creating figures with a clean and consistent look. To create publication-ready figures, we recommend that users apply additional customization based on their needs (see section [Resources](#resources)]. 


*  *  *  *


# Plot Theme {#plot_theme}

`Rgemini::plot_theme()` adds a simple theme to any `ggplot2` figures. The theme is based on `ggthemes::theme_foundation()`, with some additional tweaks, and users can easily adjust the theme as needed. By default, `plot_theme()` is applied to all other plotting functions in `Rgemini` (e.g., `plot_summary()` and `plot_over_time()`). 

The following example illustrates the theme applied to a basic barplot. We also use a color palette from `gemini_colors()` here for illustration purposes (see section [Color Palettes](#gemini_colors) for more details):

```{r}
library(ggplot2)

my_plot <- ggplot(data = data.frame(group = c("A", "B", "C"), mean = c(5, 7, 4)), aes(x = group, y = mean, fill = group)) +
  geom_bar(stat = "summary", fill = gemini_colors(2)[1:3]) +
  scale_y_continuous(name = "Mean", expand = c(0, 0))

my_plot + ggtitle("Original plot")

my_plot + 
  plot_theme() + # adds Rgemini::plot_theme
  ggtitle("Plot with default GEMINI theme")
```


## Additional inputs for `plot_theme`

`plot_theme()` accepts inputs specifying the font size/type (which are passed to `ggthemes::theme_foundation()`). Users can also specify whether to show major/minor grid lines. For example:

```{r}
my_plot + 
  plot_theme(base_size = 16, base_family = "mono", show_grid = c("major","minor")) +
  ggtitle("Mono plot with grid lines")
```


Additionally, `plot_theme()` accepts inputs that can be passed to `ggplot2::theme()`, which allows for further customization (see [here](https://ggplot2.tidyverse.org/reference/theme.html)). For example, to change the angle of the x-tick labels:
 
```{r}
my_plot + 
  plot_theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

## Overwriting `plot_theme`

If you are specifying inputs for settings that are already defined by `plot_theme()`, you will see an error message saying that the argument was matched by multiple inputs. For example, `plot_theme()` already specifies default characteristics for `plot.title` (`face = "bold"`). Thus, if you try to run the following code to italicize the title, you will see an error message saying:

```{r eval = F}
my_plot + 
  plot_theme(plot.title = element_text(face = "italic")) +
  ggtitle("Plot with additional theme")
```

`r sprintf(paste0("<span style='color: #c5070c; font-family: Lucida Console'>Error in theme(plot.title = element_text(face = 'bold', size = rel(1),  : formal argument 'plot.title' matched by multiple actual arguments</span>"))`



To avoid this error, you need to add a separate `ggplot2::theme`, which will overwrite the default characteristics specified by `plot_theme()`:

```{r}
my_plot + 
  plot_theme() +
  theme(plot.title = element_text(face = "italic"), aspect.ratio = 0.75) +
  ggtitle("Wide plot with itlicized title")
```


 
*  *  *  *

# GEMINI Color Palettes {#gemini_colors}

`Rgemini` includes a few custom color palettes that can be applied to any figures. To view all color palettes currently included in `Rgemini`, you can run:

```{r fig.width = 7, fig.height = 3}
plot_color_palettes()
```

## Colorblind Accessibility

All GEMINI color palettes are designed to be colorblind accessible. However, within each color palette, certain combinations of colors may be easier to distinguish than others. If you only need a small subset of colors, we recommend choosing the ones that are most easily distinguishable. You can check how colorblind friendly the chosen color combinations are using the `colorBlindness` R library or this website: https://www.color-blindness.com/coblis-color-blindness-simulator/



## Selecting GEMINI palettes/individual colors

All hex values for an individual palette can be obtained by running `gemini_colors()` with a numeric input (e.g., `gemini_colors(4)` for "Lavender Lagoon") or character input that matches the starting character(s) of a given palette (e.g., `gemini_colors("lav")` for "Lavender Lagoon").

For example, to create scatter plot with the colors from "Lavender Lagoon" (color palette 4):

```{r}
set.seed(2)
data <- data.frame(x = rnorm(50), 
                   y = rnorm(50), 
                   group = sample(c("A", "B", "C", "D", "E"), 50, replace = TRUE))

ggplot(data, aes(x = x, y = y, color = group)) +
  geom_point() +
  plot_theme() +
  scale_color_manual(values = gemini_colors(4)) # equivalent to gemini_colors("lav")

```

This approach works for figures generated with any R package.

## Applying GEMINI colors to `ggplots`

For figures created with `ggplot2`, you can also use `scale_fill_gemini()` or `scale_color_gemini()` to apply the color palettes. For example:

```{r}
ggplot(data, aes(x = x, y = y, color = group)) +
  geom_point() +
  plot_theme() +
  scale_color_gemini("lav")
```


## Color palettes from other packages

In addition to the GEMINI color palettes, there are several other packages that create beautiful color palettes for publication-ready figures. Below is a short intro on some of the colorblind friendly options we recommend using. These palettes allow for additional flexibility and customization that is currently not supported by the GEMINI color palettes.

### `viridis`

The viridis color palettes are a set of perceptually uniform color maps designed for scientific visualization. They were developed to address limitations of traditional color maps, which often suffer from uneven brightness or discontinuities in hue.

For more details, see [here](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html)

For example, this is a selection of 7 colors from the "viridis" palette:  

```{r fig.height = 0.5}
library(viridis)

fig <- ggplot(data.frame(x = as.factor(seq(7, 1))), aes(x = 1, y = 1, fill = x)) +
  geom_col(position = position_stack(reverse = TRUE), show.legend = FALSE) +
  coord_flip() +
  theme_void()

fig + scale_fill_viridis(discrete = TRUE, option = "viridis")
```

... or viridis' "magma" palette:  

```{r fig.height = 0.5}
fig + scale_fill_viridis(discrete = TRUE, option = "magma")
```

### `ggsci`

The `ggsci` package provides a collection of color palettes specifically designed for scientific journals. These color palettes are designed to enhance the visual appeal and interpretability of ggplot2 graphics. 

For more details, see [here](https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html and https://nanx.me/ggsci/articles/ggsci.html).

Some examples of commonly used palettes include:

+ JAMA palette:

```{r fig.height = 0.5}
library(ggsci)

fig + scale_fill_jama()
```

+ Lancet palette:

```{r fig.height = 0.5}
fig + scale_fill_lancet() 
```

+ JCO palette:

```{r fig.height = 0.5}
fig + scale_fill_jco()
```


### `RColorBrewer`

ColorBrewer is a flexible tool for selecting color schemes. ColorBrewer includes color schemes that are colorblind safe, print friendly, and perceptually uniform. It offers a wide range of color palettes optimized for different types of data and visualization purposes.

For more details, see [here](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)


```{r fig.height = 0.5}
library(RColorBrewer)
fig + scale_fill_brewer(palette = "PRGn")
```

You can explore all colorblind friendly ColorBrewer palettes by running: 

```{r fig.height = 7}
display.brewer.all(colorblindFriendly = TRUE) 
```

*  *  *  *


# Resources {#resources}

- ggplot2 vignette for theme specifications: https://ggplot2.tidyverse.org/reference/theme.html
- a comprehensive guide on ggplot2 & color palettes: https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/
- ggsci color palettes for scientific journals: https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html and https://nanx.me/ggsci/articles/ggsci.html
- viridis color maps: https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
- color brewer: https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3


<!-- remove border around images -->
<style>
img {
border: 0;
}
</style>
