---
title: "Plotting Functions - Themes & Colors"
output:
  html_vignette:
    number_sections: true
    toc: true
    toc_depth: 3
    dfprint: kable

vignette: >
  %\VignetteIndexEntry{Plotting Functions - Themes & Colors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE, 
  eval = TRUE, 
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  collapse = TRUE,
  cache = FALSE, 
  tidy = FALSE,
  fig.height = 3.5, 
  fig.width = 3.5,
  fig.align = "left"
  
)

## remove all of these later...
#library(ggpubr)
#library(ggthemes)
#library(ggpubr)
#library(scales)
#source("H:/GitHub/Rgemini/R/plot_theme.R") 

options(knitr.kable.NA = '')

```


*  *  *  *

# Introduction

This vignette introduces a standardized plotting theme ([*plot_theme()*](#plot_theme)) and color palettes ([*gemini_colors()*](#gemini_colors)) that users can apply to any ggplot figures they create in their analyses. The goal of these functions is to simplify the process of creating figures with a clean and consistent look. To create publication-ready figures, we recommend that users apply additional customization based on their needs (see section [Other Resources](#resources)]. 


*  *  *  *


# Plot Theme {#plot_theme}

`Rgemini::plot_theme()` adds a simple theme to any `ggplot2` figures. The theme is based on `ggthemes::theme_foundation()`, with some additional tweaks, and users can easily adjust the theme as needed. By default, `Rgemini::plot_theme()` is applied to all other plotting functions in `Rgemini` (e.g., `plot_summary()` and `plot_over_time()`). 

The following example illustrates the theme applied to a basic barplot. We also use a color palette from `gemini_colors()` here for illustration purposes (see section [Color Palettes](#gemini_colors) for more details):

```{r}
library(ggplot2)

my_plot <- ggplot(data = data.frame(group = c("A", "B", "C"), mean = c(5, 7, 4)), aes(x = group, y = mean, fill = group)) +
  geom_bar(stat = "summary", fill = gemini_colors(2)[1:3]) +
  scale_y_continuous(name = "Mean", expand = c(0, 0)) 

my_plot + ggtitle("Original plot")

my_plot + 
  plot_theme() + # adds Rgemini::plot_theme
  ggtitle("Plot with default GEMINI theme")
```


## Additional inputs for `plot_theme`

`plot_theme()` accepts inputs specifying the font size/type (which are passed to `ggthemes::theme_foundation()`). Users can also specify whether to show major/minor grid lines. For example:

```{r}
my_plot + 
  plot_theme(base_size = 16, base_family = "mono", show_grid = c("major","minor")) +
  ggtitle("Mono plot with grid lines")
```


Additionally, `plot_theme()` accepts inputs that can be passed to `ggplot2::theme()`, which allows for further customization (see [here](https://ggplot2.tidyverse.org/reference/theme.html)). For example, to change the angle of the x-tick labels:
 
```{r}
my_plot + 
  plot_theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) 
```

## Overwriting `plot_theme`

If you are specifying inputs for settings that are already defined by `plot_theme()`, you will see an error message saying that the argument was matched by multiple inputs. For example, `plot_theme()` already specifies default characteristics for `plot.title` (`face = "bold"`). Thus, if you try to run the following code to italicize the title, you will see an error message saying:

```{r eval = F}
my_plot + 
  plot_theme(plot.title = element_text(face = "italic")) +
  ggtitle("Plot with additional theme")
```

`r sprintf(paste0("<span style='color: #c5070c; font-family: Lucida Console'>Error in theme(plot.title = element_text(face = 'bold', size = rel(1),  : formal argument 'plot.title' matched by multiple actual arguments</span>"))`



To avoid this error, you need to add a separate `ggplot2::theme`, which will overwrite the default characteristics specified by `plot_theme()`:

```{r}
my_plot + 
  plot_theme() +
  theme(plot.title = element_text(face = "italic"), aspect.ratio = 0.75) +
  ggtitle("Wide plot with itlicized title")
```


 
*  *  *  *

# Color Palettes {#gemini_colors}

`Rgemini` includes a few color palettes with some simple options to pick basic colors. To view all color palettes currently included in `Rgemini`, you can run:

```{r fig.width = 5, fig.height = 8}
plot_color_palettes()
```


Hex values for an individual palette can be obtained by running `gemini_colors()` with a numeric input (e.g., `gemini_colors(4)` for "Lavender Lagoon") or character input that matches the starting character(s) of a given palette (e.g., `gemini_colors("lav")` for "Lavender Lagoon").

For example, to create scatter plot with "Lavender Lagoon" (color palette 4) colors:

```{r}
data <- data.frame(x = rnorm(50), 
                   y = rnorm(50), 
                   group = sample(c("A", "B", "C", "D", "E"), 50, replace = TRUE))

ggplot(data, aes(x = x, y = y, color = group)) +
  geom_point() +
  plot_theme() +
  scale_color_manual(values = gemini_colors(4)) # equivalent to gemini_colors("l")

```

If you only need a small subset of colors, we recommend using a subset of colors from a given palette that's most easily distinguishable. For example, to pick 3 equidistant colors from the "Viridis" palette (generated with `viridis`):

```{r}
ggplot(data[data$group %in% c("A", "B", "C"), ], aes(x = x, y = y, color = group)) +
  geom_point() +
  plot_theme() +
  scale_color_manual(values = gemini_colors(8)[seq(1, 9, by = 4)])
```


*  *  *  *


# Other Resources {#resources}

- ggplot2 vignette for theme specifications: https://ggplot2.tidyverse.org/reference/theme.html
- a comprehensive guide on ggplot2 & color palettes: https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/
- ggsci color palettes for scientific journals: https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html and https://nanx.me/ggsci/articles/ggsci.html
- viridis color maps: https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
- color brewer: https://r-graph-gallery.com/38-rcolorbrewers-palettes.html


<!-- remove border around images -->
<style>
img {
border: 0;
}
</style>
