---
title: "Introduction to statistical analyses with GEMINI data"
output:
  html_vignette:
    number_sections: true
    toc: true
    dfprint: kable

vignette: >
  %\VignetteIndexEntry{Statistics - Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(data.table)
library(dplyr)
library(Rgemini)
library(lubridate)

```


*  *  *  *


# Introduction & overview

This vignette illustrates an example analysis with (dummy) GEMINI data. The goal of this vignette is to provide an educational resource for analysts to make informed decisions at each analysis step. Throughout the vignette, we also provide links to existing resources with more detailed explanations.

If you would like to discuss any topics in more detail, please use the [Rgemini discussion forum](https://github.com/GEMINI-Medicine/Rgemini/discussions/categories/statistics)


*  *  *  *

# Research question & conceptual model 

- Research question: What is the effect of patients' age on clinical outcomes (mortality/LOS) among patients with community acquired pneumonia discharged between 2018 - 2020?

- Study design: retrospective cohort study -> link to epidemiology resources for other designs (e.g., cross-sectional? case control?)
- Give examples of GEMINI projects for different designs

- Bijun

## DAG 

- Understand and visualize conceptual model (confirm that project proposal/analysis plan is sound and makes sense)
  - Draw DAG
  - Define each component (e.g., what are confounders?), link to resources for mediators
  - Discuss causality
  
- Consider relevant covariates/confounders/mediators -> explain rationale for adjusted vs. unadjusted estimates
- Covariates to include here (explain variables):
  - gender
  - admission season
  - Charlson comorbidity
  - mLAPS
  - admitted from ER
  - ...

  - 1 person to start with sekelton (Kayley)
  

```{r}

```

- Additional things to consider:
  - clustering by hospital?
  - additional confounders?
  - cohort considerations, additional exclusions/inclusions required?
    - e.g., GIM vs. all-med? ER entry... ? previous hospitalizations
    - exclude COVID patients (because study period starts before COVID -> don't want results to be confounded by COVID)



## Cohort definition

Based on the DAG above, the following exclusion/inclusion criteria will be applied below.
- Incl. 1: Hospitalizations with Pneumonia as most responsible diagnosis code (MRDx) ICD-10-CA code for pneumonia in in-patient diagnosis table (need to be provided by SME); Exclude post-admission pneumonia diagnoses???
- Incl. 2: Hospitalizations with available lab data (mLAPS)
- Incl. 3: Hospitalizations admitted via ED (to exclude hospital-acquired pneumonia)
- Excl. 1: Patients with hospitalization in previous 30 day period (to exclude hospital-acquired pneumonia)

-> rationale for exclusion/inclusion criteria needs to be clear here for section below, add more explanation

-> in collaboration with PI, multiple iterations based clinical knowledge/data checks etc. (e.g., discuss during kick-off meeting)


*  *  *  *

# Data checks & cohort generation

Now that we have clarified the conceptual framework and cohort definition, let's take a look at the data.

Here, we assume our baseline cohort consists of all GEMINI encounters from 32 hospitals and we'll load this dummy data from a saved file.

```{r}
# read sample data
data <- readRDS("R:/GEMINI/Research Projects/Rgemini/stats_vignette/dummy_data_stats.rds") %>% data.table()
data <- data[ymd_hm(discharge_date_time) >= "2018-04-01 00:00" & ymd_hm(discharge_date_time) <= "2020-03-31 23:59", ]
```


We will start with some data quality/coverage checks and cohort generation. The steps in this section are not strictly sequential! Instead, data checks and cohort refinement typically go hand in hand based on an iterative process. Specifically, you will generally start with some basic checks/exploration on the whole dataset, then apply cohort inclusion/exclusion steps, then check the data for cohort-specific issues, which may cause you to refine your inclusions/exclusions etc. etc. ... and you may repeat this process several times before finalizing your cohort. Throughout that process, you will also derive additional variables/create relevant flags which should be included in your data checks to make sure all variables in your final cohort are ready to be analyzed.


## Check data coverage

- don't use term "missing" here (clarify below)

First, we'll check data coverage. Generally, this should be performed on the whole dataset within the relevant cohort period (2018-2020), prior to applying any additional inclusion/exclusion steps.

We typically check coverage by plotting variables by hospital * discharge month (because the data are pulled based on patients' discharge dates).

**These plots can provide important insights into:**
1) The number of hospitals that generally have data available for each time period
2) Some hospitals might have a very large/small number of encounters overall, which can provide information about overall patient volume (e.g., small community hospitals will generally contribute fewer encounters to the overall cohort than large academic hospitals)
3) We might observe drastic changes in variables over time (which could reflect changes in the GEMINI cohort, bugs in the derived variables, or potential biases introduced during cohort generation)

A basic sanity check is to plot the number of encounters per hospital per month:

```{r fig.width=12}
Rgemini::plot_over_time(data, func = "n")
```











Note that even if a hospital has a large number of encounters during certain time periods, this does not necessarily mean that GEMINI has all data for these encounters. This is particularly relevant for clinical data (e.g., lab/pharmacy/vitals/transfusions etc.), which may have low coverage for certain hospitals and time periods. Therefore, we should also check data coverage for any relevant clinical variables. 

For example, in our analyses, we want to include mLAPS as a covariate. Since mLAPS relies on lab data, it's possible there are certain hospitals/time periods without mLAPS values. This is illustrated in the following plot (note: `plot_over_time()` will also print a warning with all hospital * month combinations that have missing data):

```{r fig.width=12}
Rgemini::plot_over_time(data, plot_var = "mlaps") # note: you can also specify func = "na" to directly plot % missing entries
```


**Things analysts should ask themselves when looking at these plots:**
- Are there any hospitals/time periods where no lab data are available (even though administrative data may be available)?
  - E.g., hospitals 102/121 have no mLAPS values, 123/124 only have mLAPS values starting in 2019 etc.
- Are there any hospitals/time periods with implausible values/outliers?


## Apply cohort inclusions/exclusions

Next, we filter the cohort for all relevant encounters based on the inclusion/exclusion steps listed above. 

To facilitate this process, let's derive a flag for each inclusion/exclusion criteria for each `genc_id`.


### Incl. 1: Pneumonia diagnosis

When filtering by ICD-10-CA diagnosis codes, please consider the following questions:

**1. What diagnosis codes to include/exclude?** 
This is usually determined by an SME who will provide you with a list of relevant ICD-10-CA codes. In our example, all ICD-10-CA codes from J10-J18 should be included for pneumonia. Usually, you should filter for all diagnosis codes that **start** with the same characters (i.e., including children codes like "J12.8"), unless otherwise specified. If you are unsure, please clarify with the SME. 

**2. Is the diagnosis type relevant?** 
In our example, the cohort definition specifies "Pneumonia as a most responsible diagnosis code (MRDx)". We therefore need to filter by `diagnosis_type = M`. According to (HSMR methodology)[https://www.cihi.ca/sites/default/files/document/hospital-standardized-mortality-ratio-meth-notes-en.pdf] `diagnosis_type = 6` (Proxy MRDx) should also be included and takes priority over type-M.

**3. Do you want to include diagnosis codes from the ED and/or in-patient diagnosis table?**  
When filtering for a specific condition based on patient's MRDx, we typically only include in-patient diagnoses as they tend to be more reliable compared to ED diagnoses. However, this decision may be project specific, and in certain cases it might make sense to consider ED diagnoses to increase sensitivity. If you are unsure, please discuss this with the SME. 


The example data was already prefiltered for in-patient MRDx codes, so here we simply filter for the relevant diagnosis codes that start with J10-J18:

```{r}
# change dummy data to include MRDx ipdiagnosis codes!!!
# For now, add dummy diagnoses
data[, ipdiagnosis_mrdx := dummy_diag(nrow(data), nrow(data), diagnosis_type = "M", pattern = "^E11|^J1|^M23")$diagnosis_code]

data[, incl1_diag := grepl("^J12|^J13|^J14|^J15|^J16|^J17|^J18", ipdiagnosis_mrdx)]
```

-> Exclude post-admission pneumonia diagnoses???


### Incl. 2: Hospitalizations with available lab data

Since we want to include `mLAPS` as a covariate, we need to make sure lab data were in principle available, allowing us to estimate each encounters illness severity based on lab results
```{r}
#data[, incl2_lab := ]
```


### Incl. 3: Hospitalizations admitted via ED (to exclude hospital-acquired pneumonia)

```{r}
data[, incl3_ed := admitted_from_er]
```


### Excl. 1: Patients with hospitalization in previous 30 day period (to exclude hospital-acquired pneumonia)

Note: We need to calculate this based on all available data, before we apply additional exclusions
- Based on epicare???

```{r}
data[, excl1_30day := Rgemini::hospitalizations_last_n_days(data)$n_hospitalizations > 0]
```


We can now build our cohort based on the inclusion/exclusion flags as follows:


```{r}

cohort_creation_table <- function(cohort, stepnames = NULL, stepnames_short = NULL) {
  
  # get exclusion steps (will be shown as -n (-X%) to show what number/% of patients was excluded)
  excl <- grepl('^ex', stepnames_short, ignore.case = TRUE) 
  
  steps <- sapply(cohort, nrow)
  
  # create table
  tabs <- data.table(steps, lag(steps))
  tabs[, percent.change := round(1000*steps/V2)/10] 
  
  # for exclusion steps, show removal as -n (-X%) 
  if (sum(excl) > 0) {
    tabs[seq(sum(excl==F)+1, nrow(tabs)), percent.change := -1*round(1000*(1-steps/V2))/10] 
    tabs[seq(sum(excl==F)+1, nrow(tabs)), steps := -(V2-steps)] 
  }
  tabs[is.na(tabs)] <- ""
  tabs[, ':='(`N (%)` = ifelse(!is.na(percent.change), 
                               paste0(steps, " (", percent.change, "%)"), 
                               steps),
              steps = NULL, percent.change = NULL, V2 = NULL)]
  
  
  tabs <- cbind(stepnames, tabs)
  
  
  # add row with final cohort number (only needed if previous step was showing
  # exclusion
  if (excl[length(excl)] == TRUE) {
    tabs <- rbind(tabs, data.table(stepnames = "Final cohort",
                                   `N (%)` = nrow(cohort[[length(cohort)]])))
    stepnames_short <- c(stepnames_short, " ")
  }
  
  tabs <- cbind(stepnames_short, tabs)
  
  colnames(tabs) <- c("", "Cohort creation step", "N (%)")
  
  return(tabs)
}

cohort_table <- cohort_creation_table(
  cohort = list(data, # baseline cohort (discharges between 2018 - 2020) 
                data[incl1_diag == TRUE], 
                #
                data[incl1_diag == TRUE & incl3_ed == TRUE],
                data[incl1_diag == TRUE & incl3_ed == TRUE & excl1_30day == FALSE]),
  stepnames = c("All GEMINI encounters discharged between Apr 2018 - March 2020", 
                "Pneumonia MRDx", 
                "Admitted from ED",
                "Hospitalization in previous 30 day period"),
  stepnames_short = c("", "Incl. 1", "Incl. 2", "Excl. 1")
)

print(cohort_table)


```










As a sanity check, let's check what % of encounters from the original cohort this corresponds to:

```{r}
100*nrow(cohort)/nrow(data)
```
- To add: Filter by relevant time periods with available lab data (+ other inclusion/exclusion steps)


## Create additional derived variables

In the example dummy data, some derived variables have already been added (e.g., derived scores for Charlson Comorbidity, mLAPS, LOS, in-hospital mortality, CCSR condition for most responsible diagnosis code MRDx). 

Let's derive some additional covariates, like "admission season":

```{r}
data[, admission_season := Rgemini::season(admission_date_time)]
```


## Check variable distributions & missingness

Let's plot some basic descriptives and check the distribution & missingness of all relevant predictors/covariates and outcome variables:

```{r fig.height = 8, fig.width = 8}
Rgemini::plot_summary(cohort[, .(age, gender, mlaps, admit_charlson_derived, admitted_from_er, los_days_derived, in_hospital_mortality_derived, admission_season)])
```

**Things analysts should ask themselves when looking at these plots:**
- What % of entries are missing for each variable?
  - In our example, only mLAPS and Charlson Score have missing values
- What are possible reasons for missing entries?
  - mLAPS relies on lab entries, so one reason for missing entries might be that lab data are not available for certain hospitals/time periods (see below)
  - For CCI codes: In-patient diagnoses are generally available for all encounters, but diagnoses at admission may not be available if an encounter was not admitted via ED
- For continuous variables: Does the variable look normally distributed, or is it heavily skewed?
  - For example, the distribution of length of stay is heavily right-skewed (-> what to do?, links to existing resources)
- Do the frequencies/distributions generally line up with my knowledge about GEMINI data?
  - For example: Admission rate from ER is > 90%. This makes sense since GEMINI data largely represent patients with a General Internal Medicine touchpoint (or other medical subspecialties), the vast majority of whom are admitted from the ER
  
## Impute missing values

Difference between data coverage vs. missing values within time periods that have data coverage

- mLAPS: Impute missing values with 0 (if lab data generally available)
- CCI: with 0 


## Plot variables over time & by hospital (again)

... to make sure none of our inclusion/exclusion steps caused any potential biases
E.g., let's say we would want to analyze trends over time: We would want to make sure we are including the same time period from all hospitals

```{r fig.width=12}
Rgemini::plot_over_time(data, plot_var = "mlaps") # note: you can also specify func = "na" to directly plot % missing entries
```

-> adjust cohort again if necessary/perform imputation



- checks:
  - data coverage
  - missing values
  - data types
  - derived variables (categorical sex)
  - outliers
  


*  *  *  *

# Descriptive statistics

- Generate table1 and descriptive plots (but no outcome peeking) 
- Double check that numbers match up across all variables
- Confirm that our cohort has sample size to support the analysis. 

```{r}

```


*  *  *  *

# Regression model

- Fit a-priori regression models (informed by question & variable types -> decision tree illustrating different variable types and appropriate models)
- Verify regression model assumptions are met 
- Link to resources on hypothesis testing (e.g., how to interpret p-values correctly, correcting for multiple comparisons, p-hacking etc.)
- center/standardize predictors?

```{r}

```

- Do whatever other analyses (subgroups?) 
- Wherever you can, functionalize your code if you are doing the same thing over and over again 

```{r}

```


*  *  *  *

# Result visualization & interpretation

- Identify important plots, make those plots 
- Think about uncertainty (do my estimates have measure of uncertainty? If not, find way to calculate SE.)
- Check results against my intuition about what they should be, spot check randomly like this 
- Incorporate results back into the clinician’s conceptual framework so I can explain the results in their language and emphasize things that matter to them, and downplay things they don’t need to understand [iterative] 

```{r}

```


*  *  *  *

# References & resources

- https://bookdown.org/rwnahhas/RMPH/



